---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Fei Wu"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae) # For ANOVA and Tukey's HSD test
library(lubridate) # For handling date objects
library(here)
here()

DataLake <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)

DataLake$sampledate <- as.Date(DataLake$sampledate, format = "%m/%d/%y")


#2

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Lake temperature recorded during July does not change with depth across all lakes. (alpha=0)
Ha: Lake temperature recorded during July does change with depth across all lakes. (alpha does not = 0)


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

july_data <- DataLake%>%
  filter(month(sampledate) == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()
#5

ggplot(july_data, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  scale_y_continuous(limits = c(0, 35)) +
  labs(x = "Depth (m)", y = "Temperature (°C)") +
  theme_light()
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: It shows that the deeper it goes, the lower the temperature. Yes it does suggest about a linearity of this trend. 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
lm_result <- lm(temperature_C ~ depth, data = july_data)
summary(lm_result)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:

Every 1 m deeper it goes, the temperature goes down for 1.94 degree. Degree of freedom is based on the number of observations which is 9726. the p-value is smaller than 0.05 therefore it is statistically significant.73.87% of the variability in temperature is explained by changes in depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

library(corrplot)

Jul.subset <- july_data %>% 
  select(year4, daynum, depth, temperature_C) %>%
   na.omit()
corr.NTL_LTER_Jul <- cor(Jul.subset) 
corrplot(corr.NTL_LTER_Jul, method = "ellipse")
corrplot.mixed(corr.NTL_LTER_Jul, upper = "ellipse")


#10

multiregression_lake <- lm(july_data$temperature_C ~ july_data$depth + july_data$daynum)

summary(multiregression_lake)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
74% of the variability is explained by both depth and daynum. so R2 is higher now. It is an improvement. one unit of daynum, the temperature will be 0.039 degree higher. The p value is smaller than 0.05 and this is statiscally significant. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

anova_model <- aov(temperature_C ~ lakename, data = july_data)
summary(anova_model)

lm_model <- lm(temperature_C ~ lakename, data = july_data)
summary(lm_model)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 
The F-statistic and its associated p-value (< 2.2e-16) from the summary indicate that the model as a whole is statistically significant. This means there is strong evidence to reject the null hypothesis that there is no difference in mean temperatures across the lakes, which means that the mean temperature varies significantly among the lakes.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

ggplot(july_data, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(limits = c(0, 35)) +
  labs(x = "Depth (m)", y = "Temperature (°C)", color = "Lake Name") +
  theme_light()
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
tukey_result <- TukeyHSD(anova_model)
tukey_result

lake_groups<- HSD.test(anova_model, "lakename", group=TRUE)
lake_groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: If we look at group c, Paul Lake and Ward Lakehave the same mean temperature as Peter Lake.No single lake has a mean temperature that is statistically distinct from all the others in a way that it forms its own unique group without overlap. 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: assume the data follows a normal distribution, a t-test would determine if there are statistically significant differences between their mean temperatures.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}


DataLake2 <- july_data %>%
filter(lakename%in% c("Crampton Lake", "Ward Lake"))
t_test_result <- t.test(temperature_C ~ lakename, data = DataLake2)
t_test_result
```

>Answer: . The p-value of 0.2649 is greater than 0.05, suggesting that we fail to reject the null hypothesis that there is no difference in the mean temperatures.The mean for the two lake temperatures are not equal, but they are not significantly different from each other either. This proves no.16 that no lake temperature is significantly distinct from one another. 
