---
title: "Assignment 3: Data Exploration"
author: "Fei Wu"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}

#install.packages("tidyverse")
library(tidyverse)

Neonics <- read.csv("./Data/Raw/Neonics.csv",stringsAsFactors=TRUE)
Litter <- read.csv("./Data/Raw/NIWO_Litter/Litter.csv", stringsAsFactors = TRUE)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Key reasons for interest include concerns about the impact on non-target organisms, particularly pollinators like bees, potential effects on biodiversity and ecosystems, water contamination risks, regulatory considerations for pesticide use, and potential indirect effects on human health through the food chain. Understanding the ecotoxicological effects informs regulatory decisions and aids in mitigating harm to both ecosystems and human well-being.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Fallen debris plays a key role in nutrient cycling, providing habitats for diverse organisms, and influencing carbon sequestration. Monitoring litter and woody debris at Niwot Ridge contributes valuable insights into long-term ecosystem health, helping assess the impact of climate, disturbance, and land use changes. Additionally, it aids in fire ecology research, as the debris can act as both fuel and a barrier. Overall, this research enhances our understanding of how forest ecosystems function, adapt, and respond to environmental changes, providing essential information for effective ecosystem management and conservation strategies.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Ground taps are sampled once a year
 2.Litter and fine woody debrits are collected from elevated and ground traps respectively.
 3. Litter is defined as material that is dropped from the forest canopy and has a butt end diameter <2cm and a length<50cm

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
Neonics.data.complete <- na.omit(Neonics)
dim(Neonics)
dim(Neonics.data.complete)

## Dimensions is 4623,30
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
summary(Neonics$Effect)
```

> Answer: Mortality and Population is the most common effectss. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r}
summary(Neonics$Species.Common.Name)

#Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, Italian Honeybee. 
```


8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author)
```

> Answer:It's factor becasue they not just all numbers and we read them as strings earlier.
## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 50)

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}

ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color=Test.Location), bins = 50) 

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Lab. they do differ over time. Between 1990 to 2000, field natural is low but becomes high in 2010. Field artificial did not appear until late 2000 to 2010. Lab was always there and peaked between 2010 to 2020. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint), bins = 50) + theme(axis.text= element_text(size=10, angle=90))

```

> Answer:LOEL and NOEL

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate)
Litter$collectDate <-as.Date(Litter$collectDate)
class(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID)
summary(Litter$plotID)
```

> Answer: levels means lising out every unique value ; summary will tell us the unique values and the number of them appear 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
ggplot(Litter) +
  geom_bar(aes(x = functionalGroup), bins = 50)

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y=dryMass), bins = 50)

ggplot(Litter)+
  geom_violin(aes(x=functionalGroup, y=dryMass))
  ```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> #Answer:not enough values for violin.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: #Needles 
