---
title: "Assignment 2: Coding Basics"
author: "Fei Wu"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
seq (1,30,3)
thirty_sequence <-seq(1,30,3)
# I used the seq function to determine the numbers of the sequence and I named it as "thirty_sequence".

#2. 
mean(1,4,7,10,13,16,19,22,25,28)
median(1,4,7,10,13,16,19,22,25,28)
# i copied and pasted what R gave me in terms of the numbers and used the mean and median function to determine the mean and median of the sequence.
#3. 
1>1
# I used what R gave me for mean and median, and I asked R if mean is bigger than the median. R gave me the answer "False".
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
# 5
names_vector <- c("student1","student2", "student3", "student 4")
# student names vector; strings
test_score_vector <- c(100, 80,45,50)
# test score vector; numerical
pass_vextor<- c(TRUE, TRUE,FALSE,TRUE)
# pass or not vector; logial items

# 7
students_performance <- data.frame(names_vector, test_score_vector, pass_vextor)
students_performance
# 8
create_df <- data.frame("Student Names"=names_vector,"Test Score"=test_score_vector, "Passed"= pass_vextor)
create_df
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:matrix can only contain a single class of data, while this data frame can consist of many different classes of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
create_df$Test.Score
Function_1 <-function(Test_score){
  ifelse (Test_score>50,TRUE,FALSE)}

Function_1(create_df$Test.Score)
      
  ```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Becasue "if" and "else" only works on single value.
